---
title: "test 1D 3Cat HMNL sum to zero"
output: html_notebook
---

# Sum to zero coefficient HMNL 1 dimensional with lapses optional

This is a demonstration of Bayesian hierarichical multinomial regression with lapses in responses run on some test data.

```{r}
library(bhmnl)

system(paste0("open ",  "'"  ,  tempdir() ,  "'" )) 

# Data generated thru "CreateTest2D3Cat_HMNL_data.Rmd"  in this dir
useLapse=TRUE
if(useLapse){
  browser()
  load("../data/test1D3CatLapseDatList.rda")
srDtbAll= test1D3CatLapseDatList$srDtbAll
thetaMatTrue=test1D3CatLapseDatList$thetaMatTrue
bMatSjList= test1D3CatLapseDatList$bMatSjList
lambda_sList=test1D3CatLapseDatList$lambda_sList

rm(test1D3CatLapseDatList)
}else{
data("test1D3CatDatList.rda")
# Peel out the list elements
srDtbAll= test1D3CatDatList$srDtbAll
thetaMatTrue=test1D3CatDatList$thetaMatTrue
bMatSjList= test1D3CatDatList$ bMatSjList
rm(test1D3CatDatList)
}
# Response factor:

srDtbAll[,resp:=factor(catSym,levels=c("A","B","C"))]
# 
formAna= ~resp*x

rsSumDmx=make_rsSumDmxFrom_srDtb(form = formAna, srDtb = srDtbAll,
                                 stimVarColNames = c("x"), respFacDescr = c("resp"),
                                 includeMasterCatInDesired = TRUE)


```
```{r}
str(rsSumDmx)
masterCatLevels=attr(rsSumDmx,'masterCatLevels')
nCat=length(masterCatLevels)

```
##  Now lets try BayesHMNL with sum to zero coefficients
When an element is of type list, it is supposed to make it easier to pass data
for those declared in Stan code such as 
"vector[J] y1[I]" and "matrix[J,K] y2[I]".
Using the latter as an example, we can use a list for y2 if the list has "I" elements,
each of which is an array (matrix) of dimension "J*K". However, it is not possible
to pass a list for data declared such as "vector[K] y3[I,J]"; the only way for it is to
use an array with dimension "I*J*K".
variable changes nRespCat -> nRespCat ; nP -> P participant; T -> T total number of observation trials
where each observation  trial is a single trial for a single participant.

```
data {
int<lower=2> nRespCat; // Number of response category alternatives (choices) in each scenario
int<lower=1> K; // Number of columns in each design matrix list X[t]
int<lower=1> nSj; // Number of subjects (respondants/ agents/ persons/perceivers)
int <lower=1> nT; // total number of observation trials  (grand trials) ////
int<lower=1,upper=nRespCat> Y[nT]; // YB[nSj, S]; // best choices
matrix[nRespCat, K] X[nT]; // was   matrix[nRespCat, K] X[nSj, S]; // matrix of attributes for each obs
int<lower=1, upper=nSj> SjID[nT]; // Added tmn. Serial identifier for participant on each observation trial
}

```
THis means we need to pack 
rsSumDmx into a huge list
with nrows(rsSumDmx)/nCat elements
```{r}

nT=nrow(rsSumDmx)/nCat
# nRep=nT/nStim*nSj
#  assertthat::assert_that(nT==nStim*nSj*nRep) 
XList=list()
irng=1:nCat
# Repack the data
for (t in 1:nT){
  XList[[t]]=as.matrix(rsSumDmx[irng,])
  irng=irng+nCat
}
```
# Run the sucker
parameters {
// vector[K] Beta[nSj];
vector[K]  Beta[nSj];
// vector[K - 1] Theta_raw;
vector[K] Theta;
cholesky_factor_corr[K] L_Omega;
vector<lower=0, upper=pi()/2>[K] L_sigma_unif;
}
#  Pick analysis

```{r}
# Fully non-centered version or not
if (useLapse){
  # stanFile=  "../inst/stanFiles/BayesHMNL_s2zLapseM1A.stan"
     stanFile= system.file("stanFiles","BayesHMNL_s2zLapseM1A.stan", package="bhmnl")

  saveParsCvec=c('theta','beta','L_sigma','L_Omega', 'u')
}else{
  catln('BayesHMNL_s2zM1A.stan' )
  # stanFile=  "../inst/stanFiles/BayesHMNL_s2zM1A.stan"
     stanFile= system.file("stanFiles","BayesHMNL_s2zM1A.stan", package="bhmnl")
  saveParsCvec=c('theta','beta','L_sigma','L_Omega')
}

homogeneous=FALSE
if (homogeneous){
  ### THIS WORKS FINE BELOW
  ## Simple homogeneous logit see if we've got it right
  # stanFile=  "../inst/stanFiles/BayesMNL_s2z.stan"
  stanFile= system.file("stanFiles","BayesMNL_s2z.stan", package="bhmnl")

  saveParsCvec=c('theta')
}
### AV

library(rstan)
nChains=4
nIter=2000
nWarmup=floor(2*nIter/4)
nRefresh= 100
myVerbose= FALSE
testData=list(X=XList,
              Y=as.integer(srDtbAll$resp),
              jj= as.integer(srDtbAll$Sj),
              D=ncol(XList[[1]]),
              C= nCat,
              N=nT,
              J=length(unique(srDtbAll$Sj)),
              dbgLev=0
)
```

## Remove unnecessary items from memory
```{r}
rm(list=c('srDtbAll','rsSumDmx'))

```

```{r}
fit1NC <- stan(
  file = stanFile,  # Stan program
  data = testData,    # named list of data
  pars = saveParsCvec,
  chains = nChains,             # number of Markov chains
  warmup = nWarmup,          # number of warmup iterations per chain
  iter = nIter,            # total number of iterations per chain
  cores = 2,              # number of cores (could use one per chain)
  control=list(adapt_delta=.95),
  refresh = nRefresh,             # progress every nRefresh
  verbose = myVerbose,
  init="0" # This was running off the rails with random initialization
  
)

```


```{r}
show(fit1NC)
# summary(fit1NC)
# sampler_params <- get_sampler_params(fit1NC, inc_warmup = TRUE)
# summary(do.call(rbind, sampler_params), digits = 2)
# stop('Stop here')

```

# Compare to  theta generating data -- the non-lapse true theta values
We see that estimated 'population' mean is very close to true bSj= betaSj+theta values in the generating sample. (The third column in the true means is redundant -- sum to zero across cols and is not estimated)

```{r}
# https://mc-stan.org/rstan/reference/stanfit-method-extract.html
fit_ss <- extract(fit1NC, permuted = TRUE) # fit_ss is a list 
thetaH=fit_ss$theta
thetaHMeanVec=apply(thetaH,2,mean)

# 
# thetaMatTrue
#            [,1]     [,2]       [,3]
# [1,] -0.5270463 1.054093 -0.5270463
# [2,] -4.0000000 0.000000  4.0000000
# We only want to compare cols 1 and 2
# 
 # https://stackoverflow.com/a/16543224/1795127
 
thetaTrueVec=as.vector(t(thetaMatTrue[,-3]))
print(thetaTrueVec)
bArraySj= simplify2array(bMatSjList)

bArraySjMean=apply(bArraySj,c(1,2),mean)
bSjMeanTrueVec=as.vector(t(bArraySjMean[,-3]))

catln('thetaTrueVec')
print(thetaTrueVec)
catln('bSjMeanTrueVec')
print(bSjMeanTrueVec)
catln('thetaHatMeanVec')
print(thetaHMeanVec)

# 

```
# Compare to u -the key lapse parameters (u's) by subject
```{r}
uH=fit_ss$u
# uH
#  dim(uH) samples, subjects, parameters
# [1] 4000   10    3
# Get mean over iterations  (rows are subjects)
uHMeanSjMat=apply(uH,c(2,3),mean )

#  what you want is in 
#  > lambda_sList[[iSj]]$sU
# # # $lambda_s
# # [1] 0.001357835 0.013273951 0.016028824
# # $lambda
# # [1] 0.03066061
# # $s_u
# # [1] 0.04428598 0.43293172 0.52278230

xx=c()
yy=c()
for (iSj in 1:nrow(uHMeanSjMat)){
  catln("Subject", iSj, "uHat then uTrue one line each")
  # but our uH  need exponentiation
  print(exp(uHMeanSjMat[iSj,]))
  xx=c(xx,uHMeanSjMat[iSj,])
  print(lambda_sList[[iSj]]$s_u)
  yy=c(yy,lambda_sList[[iSj]]$s_u)
}
plot(xx,yy)
```

## Bayesplot
```{r}
library(bayesplot)
color_scheme_set('gray')
posterior_cp=as.array(fit1NC)
nutsPars <- nuts_params(fit1NC)
```
# Rhats

```{r}
# seems to require bayesplot:: prefix
myRexPars="^theta|^beta|^L_sigma|^u"
rhats=bayesplot::rhat(fit1NC,regex_pars=myRexPars)
print(rhats[rhats>1.1])

```
```{r}
# seems to require bayesplot:: prefix
rhats=bayesplot::rhat(fit1NC)
print(rhats[rhats>1.1])
mcmc_rhat(rhats,regex_pars=myRexPars)
```


## Neff
```{r}
# THis onw needs pruning
neffRatios=neff_ratio(fit1NC)
print(neffRatios)
 mcmc_neff(pruneStanfitObj(fit1NC, c("theta", "beta", "L_sigma", "u")))
```
## Parcoord


```{r}
mcmc_parcoord(posterior_cp, np = nutsPars, regex_pars = '^theta')
```
```{r}
mcmc_parcoord(posterior_cp, np = nutsPars, regex_pars = '^L_sigma')
```

```{r}
mcmc_parcoord(posterior_cp, np = nutsPars, regex_pars = '^u')
```


```{r}
mcmc_pairs(fit1NC,np = nutsPars,regex_pars="^beta\\[[1-9],2|^L_sigma\\[2]")
```

```{r}
pairs(fit1NC, pars = c("theta[1]","theta[2]", "theta[3]", "L_sigma[1]",  "L_sigma[2]",  "L_sigma[3]"), log = TRUE, las = 1)
```

```{r}
catln('About to launch shinystan')
browser()
library(shinystan)
launch_shinystan(fit1NC)

# 
# 
```

