---
title: "Factorial Response Logistic Demo"
output: html_notebook
---

# Generate "Indicator variable" redundant design matrix with parseable labels
The general strategy:   Create an "inflated factorial design" to trick `model.matrix()`
to designing an "Indicator variable" model with standard formulae.
https://stackoverflow.com/a/56677013/1795127
Reason: We want MacElreath style "Indicator variables" for each factor and interaction of a possibly factored set of responses response. 

##Outline:  

* For K-factors [1:K] each witn $l_k$ factors each (Works fine for K==1):
* Pad the levels argument of factor with an extra "0-th level" response level, (so there are $l_k +1$ levels in the inflated factor) and label  the extra level 0 as "_".
* We use `car::contr.Treatment` to get clear labeling of levels
* We then use `model.matrix()` to generate the inflated design matrix, including an unwanted "(Intercept)"  and response main effect (no interactions with response factors) terms. 
* Finally, we extract the matrix  part only, losing the spurious intercept term.

### This also handles factored stimuli. 
Factored will have internally expanded levels. The factored stimulus levels will will be created with default r contrasts with default r contrast names. <facName><lev>
Factored responses will (via car::contr.Treatment) have the form <facName>|<level>.

We also demonstrate packing data for an example stan file `./yegBaselineHMNL.stan`
(See also: [https://discourse.mc-stan.org/t/speeding-up-a-hierarchical-multinomial-logit-model/1538/5?u=tnearey]
)

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 
```{r}
library(data.table)
# actually only need car::contr.Treatment
# library(car)
# Is comment ok
catln <- function(...) {cat(...,'\n')}

```

## >> myFactorialHMNL_CreateData()
Create some test data

## Create a fully specified response-Indicator-variable model.
### Read in test data into stimulus response data.table, srDtb

### This assumes that there is a coherent factor specification to master response category (here `syl' ) and any coherent factorizations of the master. Here 'cn' and 'vow'.
#### A generalized version will follow

# prepareFacRespDmxDtb=function(masterCatVName,catFactVNames,stimVNames

## >> showRespFacDtbLevs()
```{r}
## Debugging 
showRespFacDtbLevs=function(respFacDtb){
  for (n in names(respFacDtb)){
    catln(n, levels(respFacDtb[[n]]))
  }
}
```

## >> showColInfo(xdtb) function
Show compact summmaries
```{r}
showColInfo=function(xdtb){
  catln('class of input', class(xdtb))
  for (n in names(xdtb)){
    t=xdtb[[n]]
    catln(n, class(t))
    nna=sum(is.na(t))
    nuniq=length(unique(t))
    if (is.factor(t)){
      catln(levels(t))
    }else if (is.numeric(t)){
      catln('min', min(t), 'max', max(t), 'nuniq', nuniq)
    }else{
      catln('nuniq', nuniq)
    }
    if (nna>0) {catln('*** n na ',nna, '***')}
  }
}
```
## >> checkAllAreFactors
```{r}
checkAllAreFactors=function(facNames,dtb){
  checkit=TRUE
    for (f in facNames){
    if (!is.factor(dtb[[f]])){
      catln('Not factor: ', f)
      checkit=FALSE
    }
    }
  if (!checkit) stop('Must all be factors')
  return(checkit)
}

```


## >> make_respFacDtb() 

```{r}
#' make_respFacDtb Make a response factor data tabble
#'
#' @param srDtb  -- a stimulus-response data.table (one trial per row)
#' @param masterCatColName -- column name for master category
#' @param desiredFactorColNames -- column names to be used in formula
#' @param includeMasterCatInDesired -- boolean (default TRUE) whether to include masterCatColName among desiredFactorColNames

#' @details  Collect response factors from an srDtb and collect into a compact respFacDtb 
#' Note if you want to exclude  masterCatColName from output of columns,
#' you must set includeMasterCatInDesired=FALSE
# .
#' @return - returns a respFacDtb 
#' @export
#'

make_respFacDtb=function(srDtb,masterCatColName,desiredFactorColNames, includeMasterCatInDesired=TRUE){
  
  facnames=unique(c(masterCatColName,desiredFactorColNames))
if (!checkAllAreFactors(facnames,srDtb)) stop(' Programming error')
  
  inxFirstAppearanceMasterCat=c()
  if (includeMasterCatInDesired && !(masterCatColName %in% desiredFactorColNames )){
    desiredFactorColNames=c(masterCatColName,desiredFactorColNames)
  }
  masterCatLevs=levels(srDtb[[masterCatColName]])
  for (i in seq_along(masterCatLevs)){
    s=masterCatLevs[i];
    # Get first occurrence of each syl, cn and vow corresponding to each level of syl
    inxFirstAppearanceMasterCat[i]=which(srDtb[[masterCatColName]]==s)[1]
  }
  respFacDtb=srDtb[inxFirstAppearanceMasterCat, .SD,.SDcols=desiredFactorColNames]
  
  return(respFacDtb)
}
```



This is to trick model.matrix into specifying a full matrix
# >>> padLevels() - required by padLevels_facRespDtb, expand_facStimDtbLevels
pad levels of single factor
```{r}
padLevels=function(x){ 
  y=factor(x,levels=c('_',levels(x)))
}
```
## >> padLevels_facRespDtb().  Expand the levels of a factor with a blank one at the front

```{r}
padLevels_facRespDtb= function(respFacDtb){
  # Note may does NOT work "in place" as indexing via []
  # Expand the levels with a dummy blank one at the front
  # This is to trick model.matrix into specifying a full matrix
  # This appears to work:
  # Get a simple name for car::contr.Treatment and set the decorator
  # This could all be done inside data.table by referene world, 
  # using .SD, .SDcols and .L apply,
  # I think we are triggering a copy and maybe a conversion to data.frame
  for (fac in names(respFacDtb)){
    respFacDtb[[fac]]=padLevels(respFacDtb[[fac]])
    # catln(fac, 'still data .table', data.table::is.data.table(respFacDtb))
  }
  # catln(' before return padLevels_facRespDtb')
  # showRespFacDtbLevs(respFacDtb)
  return(data.table::as.data.table(respFacDtb))
}
```
#
## >> padLevels_facStim_srDtb(srDtb,stimVarColNames)
```{r}
padLevels_facStim_srDtb= function(srDtb,stimVarColNames){
  # Expand any factorial stimulus factors
  for (fac in stimVarColNames){
    if (is.factor (srDtb[[fac]])){
      srDtb[[fac]]=padLevels(srDtb[[fac]])
      # catln(fac, 'still data .table', data.table::is.data.table(srDtb))
      # catln('levs stimfac', fac)
      # print(levels(srDtb[[fac]]))
    }
  }
  # catln(' before return padLevels_facRespDtb')
  
  return(as.data.table(srDtb))
}
```

## >> make_stimRespFacDtb(respFacDtb,srDtb)
THis constructs  a respFactDtb row expanded [responesFactors, stimFactors] NROWS (== (  nrow(srDtb) x nMasterRespCat )  ) rows
 with stimulus factors varying fastest down rows, then response categories

```{r}
make_stimRespFacDtb= function(respFacDtb,srDtb,stimVarColNames){
  # This is slowish but it's pretty small . Could be done with .SD .SDcols and lapply, but this is clearer.
  
  ### Set up the car:Treatment contrasts for clearer labeling. Then build the srDmx design matrix
  #  CBIND the columns of expanded  srDtb and respFacDtb factors, and make it a model.matrix
  # Then extract just the design part
  # paste the columns of row expanded respCat factors and row expanded stimulus properties
  # Data table is useful because we can copy its rows efficiently.
  
  nObs=nrow(srDtb)
  # Get this before expansion of factors
  nrespCat=nrow(respFacDtb)
  ## Need to keep track of stimulus properties extracted from srDtb
  inxSrDtbCopy=rep(1:nObs,times=nrespCat)
  inxRespFacCopy=rep(1:nrespCat,each=nObs)
  stimRespFacStimDtb = cbind(respFacDtb[inxRespFacCopy, ],srDtb[inxSrDtbCopy, .SD,.SDcols=stimVarColNames])
}

```
## >> make_respFacStimDtb(respFacDtb,srDtb)
THis constructs  a respFactDtb row expanded [responesFactors, stimFactors]nObs (== ( nMasterRespCat x nrow(srDtb))  ) rows
 with response factors varying fastest down rows

```{r}
make_respFacStimDtb= function(respFacDtb,srDtb,stimVarColNames){
  # This is slowish but it's pretty small . Could be done with .SD .SDcols and lapply, but this is clearer.
  
  ### Set up the car:Treatment contrasts for clearer labeling. Then build the srDmx design matrix
  #  CBIND the columns of expanded  srDtb and respFacDtb factors, and make it a model.matrix
  # Then extract just the design part
  # paste the columns of row expanded respCat factors and row expanded stimulus properties
  # Data table is useful because we can copy its rows efficiently.
  
  nObs=nrow(srDtb)
  # Get this before expansion of factors
  nrespCat=nrow(respFacDtb)
  ## Need to keep track of stimulus properties extracted from srDtb
  inxSrDtbCopy=rep(1:nObs,each=nrespCat)
  inxRespFacCopy=rep(1:nrespCat,times=nObs)
  
  
  respFacStimDtb = cbind(respFacDtb[inxRespFacCopy, ],srDtb[inxSrDtbCopy, 
                                                              .SD,.SDcols=stimVarColNames])
}
```
## >> make_Dmx(form, respFacStimDtb,respVarColNames,stimVarColNames)
```{r}
make_Dmx=function(form, respFacStimDtbPadded,respVarColNames,stimVarColNames){
  # Define contr.treat and decorator here.
  # Make the model matrix don't forget the 1.
  # Add also bare stimulus variables involved in formula
  # The intercept and stimvars will get stripped at end.
  # Note if respFacStimDtbPadded then you get an rsDnx with response cat
  # varying fastest down rows, if stimRespFacDtbPadded, then you get 
  # an srDmx with stinulus varying fastest and response cats stacked.
  if (class(form)=='formula'){
    # https://stackoverflow.com/a/14671346/1795127
    # Eliminate extra spaces too
    formStr= gsub('  +',' ',Reduce(paste, deparse(form)))
  }else if (is.character(form)){
    formStr=form
  }else{
    stop(' Input `form` should be  of class character string or formula')
  }
  # Get rid of anything up to ~ in formStr
  formStr=sub('~','',formStr)
  varsInForm=all.vars(as.formula(paste0('~',formStr)))
  # Find the stim varsvariables that are included in the formula
  stimVarsInForm = intersect(stimVarColNames, varsInForm)
  respVarsInForm = intersect(respVarColNames,varsInForm)
  # catln('stimVarsInForm',stimVarsInForm)
  # catln('respVarsInForm',respVarsInForm)
  # Also get factored stimuli
  # factoredStimInForm=stimvarsInForm(sapply(is.factor,)
  # sapply(is.factor )
  # 
# Get names of stim factors in form to add then in to mmCotrastList
binxStimFactors=sapply(stimVarsInForm, function(x) is.factor(respFacStimDtbPadded[[x]]))
stimFactorsInForm=stimVarsInForm[binxStimFactors]

mmContrastList=list()
  for (sv in c(respVarsInForm,stimFactorsInForm)){
    # Compile the contrast matrix
    mmContrastList[[sv]]="myIndicatorContrFun"
  }
  catln('mmContrastList')
  print(mmContrastList)

  # ???We have to add any stim vars to near the begining of the formula to keep from triggering "dummy" coding 
  # in model.matrix. This includes factored stimulus variables (I think) 

  nstimVarsAug= length(stimVarsInForm)
  # paste0( "1+ ", paste0(f,collapse=' + '), ' + ', 'a + b + c')
  # Paste the stimulus variables at the END?
  augFormStr=paste0(" ~ 1 + ",formStr  , 
                    " + " ,paste0(stimVarsInForm, collapse = " + ") )
  # catln('formStr' ,formStr)
  # catln('augFormStr',augFormStr )
  mmx=model.matrix(as.formula(augFormStr) ,respFacStimDtbPadded, contrasts.arg = mmContrastList)
  
  # catln('mmx')
  # summary(mmx)
  # catln(' COlnames mmxRef')
  # 
  mmxNames=colnames(mmx)
  # Remove the columns  exra columns 1 + number of stimvars in formula
  # Good terms hould have contrast "|l<level> in in them either at begining or after :
  # binxKeep=grepl('(^|:).*?\\|',mmxNames)
  # Above doesn't work.
  #  We need to delete any column that doesn't have a stimulus variable
  #  This will do it.
  effectsList= lapply(strsplit(mmxNames,':'), function(x) gsub('\\|.*$','',x))
  binxKeep=sapply(effectsList, function(x) any(x %in% respVarsInForm))
  # catln('broswer')
  # browser()
  # catln('passing mmx names', mmxNames[binxKeep])
  srDmx=as.matrix(mmx)[,binxKeep]
  return(srDmx)
}

```
##  >> myIndicatorContrFun()
```{r}
#' myIndicatorContrFun   a decorated contrast function always forcing INDICATOR levels
#'
#' @param levLabs  charvec of level labels
#' @param base base category
#' @param contrasts  contrast=TRUE (ignored)
#'
#' @return
#' @export
#'
#' @examples
# cat('no examples\n')
myIndicatorContrFun = function (levLabs, base = 1, contrasts = TRUE)
{
  # Specialized function for labeling contrsts witn <facname>|<levelLab>
  # I.e. pipe symbol as prefix for level,
  # inspired by car:contr.Treatment
  catln('class (levLabs) in myIndicatorContrFun', class(levLabs))
  catln('base', 'contrasts', base, contrasts)
  # readline('myIndicatorContrFun :')
  if (!is.character(levLabs)) {
    stop('levLabs (first arg) needs to be a character vector.')
  }
  n <- length(levLabs)

  if (!(base==1 && contrasts)){
    stop( 'myIndicatorContrFun must have base=1 and contrasts=TRUE')
  }
  contrastNames <- paste0('|', levLabs)
  contr <- array(0, c(n, n), list(levLabs, contrastNames))
  diag(contr) <- 1
  if (contrasts) {
    if (n < 2) 
      stop("Contrast requires at least 2 d.f.")
    if (base < 1 | base > n) 
      stop("Base level must be >= 1 and <= n")
    contr <- contr[, -base, drop = FALSE]
  }
  contr
}
```
## TestMake_respFacStimDtb
```{r}
## 
## 
# 
# ---------TestMake_respFacStimDtb= function(){ ----------
# Future expansion ??

tst=readline( "do test y/n? ")
srDtb=data.table(myFactorialHMNL_CreateData())
# str(srDtb)
# Classes ‘data.table’ and 'data.frame':	100 obs. of  6 variables:
#     $ syl   : Factor w/ 6 levels "ba","bi","bu",..: 4 3 2 6 6 1 5 6 5 2 ...
# $ x     : num  0.253 -2.029 -2.043 1.369 -0.226 ...
# $ y     : num  0.788 0.769 2.332 -1.008 -0.119 ...
# $ chosen: Factor w/ 6 levels "ba","bi","bu",..: 4 3 2 6 6 1 5 6 5 2 ...
# $ cn   : Factor w/ 2 levels "b","d": 2 1 1 2 2 1 2 2 2 1 ...
# $ vow   : Factor w/ 3 levels "a","i","u": 1 3 2 3 3 1 2 3 2 2 ...
# - attr(*, ".internal.selfref")=<externalptr
#
#


respFacDtb=make_respFacDtb(srDtb,'syl',c('syl','cn','vow'))
respVarColNames = names(respFacDtb)
catln('before respFacDtb expansion')
showRespFacDtbLevs(respFacDtb)
# readline('Here')
# 
 # ====== experimental ==============
 # no padding using myIndicatorFun for 'contrasts.arg' list
 respFacStimDtb= make_respFacStimDtb(respFacDtb,srDtb)
# Padding requried for standard contrst (eg myIndicatorContrFun, contr.Treatment...)
respFacDtbPadded=padLevels_facRespDtb(respFacDtb)

catln('after respFacDtb padding')
showRespFacDtbLevs(respFacDtbPadded)
stimVarColNames=c('x','y','m')
# Pad any factorial stimulus levels
srDtbPadded= padLevels_facStim_srDtb(srDtb, stimVarColNames)
respFacStimDtbPadded= make_respFacStimDtb(respFacDtbPadded,srDtbPadded)
stimRespFacDtbPadded= make_stimRespFacDtb(respFacDtbPadded,srDtbPadded)
showColInfo(respFacStimDtbPadded)
showColInfo(stimRespFacDtbPadded)
# Costruct a stimulus-response design matrix
```

# Execute make dmx from padded dtbs
```{r}

if (tst!='y') stop('Stopped by user')
# Maybe if this is defined before the call to make_rsDmx?
# contrTreat=function(...) car::contr.Treatment(...)
# This has responsecategory varying fastest in rows then stimuli
rsDmx= make_Dmx(~cn+vow+cn:vow+cn:x+vow:y+cn:m, respFacStimDtbPadded,respVarColNames,stimVarColNames)
srDmx=make_Dmx(~cn+vow+cn:vow+cn:x+vow:y+cn:m, stimRespFacDtbPadded,respVarColNames,stimVarColNames)
catln('rsDmx')
print(head(rsDmx,n=20))
catln('srDmx')
print(head(srDmx,n=20))
```

## Feed to RSTAN
// Feeding data. Should work if we provide data as per // https://mc-stan.org/rstan/reference/stan.html
ˆ
"matrix[J,K] y2[I]... we can use a list for y2 if the list has "I" elements,
//    each of which is an array (matrix) of dimension "J*K"
 # W need  matrix [ C, K]  X [N]
```{r}

X=list()
rowOffset=0
# We also have to pull off the subject (Respondant) variable
for (n in 1:nObs){
X[[n]]=rsDmx[rowOffset+1:C, ]
rowOffset=rowOffset+C
}

# data {
#   int<lower=2> C; // Number of alternatives (choices) in each scenario
#   int<lower=1> K; // Number of alternatives
#   int<lower=1> R; // Number of respondents
#   int <lower=1> N; // total number of observations (grand trials) //// int<lower=1> S; // Number of scenarios per respondent
#   int<lower=1,upper=C> Y[N]; // YB[R, S]; // best choices
#   // int<lower=1,upper=C> YW[R, S]; // worst choices
#   matrix[C, K] X[N]; // was   matrix[C, K] X[R, S]; // matrix of attributes for each obs
#   int<lower=1, upper=R> RID[N]; // Added tmn. Serial identifier for respondant r on grandtrial N.
}
stanData=list( N=as.integer(nObs), C= as.integer(nrespCat), K = as.integer(ncol(rsDmx)),
               R=


)
```












```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
srDmxRef


